Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-07-01 ms.tex
Some minor changes
New citations and some text about the RoI selection and updated tables

\altaffiltext{1}{Corresponding author: N. Karlsson, nkarlsson@physics.umn.edu}
\altaffiltext{2}{School of Physics and Astronomy, University of Minnesota, Minneapolis, MN 55455, USA}
\altaffiltext{3}{DESY, Platanenallee 6, 15738 Zeuthen, Germany}
\altaffiltext{4}{ Institut fu\"r Physik und Astronomie, Universita\"t Potsdam, 14476 Potsdam-Golm, Germany}
\altaffiltext{5}{Heidelberg Institute	for Theoretical	Studies, Schloss-Wolfsbrunnenweg 35, D-69118 Heidelberg, Germany}
\altaffiltext{6}{Canadian Institute for Theoretical Astrophysics, University of Toronto, Toronto, ON, M5S 3H8, Canada}
\altaffiltext{7}{University of California, Santa Barbara, CA 93106}

@@

The Coma cluster of galaxies (ACO 1656) is one of the most thouroughly studied clusters over all wavelengths \citep{article:Voges_etal:1999}. Located at a distance of about 100 Mpc \citep[$z=0.023$;][]{article:StrubleRood:1999}, it is one of the closest and also one of the most massive clusters \citep[$M \sim 10^{15}M_{\odot}$;][]{article:Smith:1983, article:Kubo_etal:2008}. It hosts both a giant radio halo \citep{article:Giovannini_etal:1993,article:Thierbach_etal:2003} and peripheral radio relic, which appears connected to the radio halo with a ``diffuse'' bridge \citep[see discussion in][]{article:BrownRudnick:2010}. It has been suggested that the relic is an infall shock. Extended soft X-ray (SXR) emission is evident from the ROSAT all-sky survey in the 0.1 to 2.4 keV band \citep{article:BrielHenryBohringer:1992} and observations with XMM-Newton \citep{article:Briel_etal:2001} revealed substructure in the X-ray halo. As such it is a natural candidate for gamma-ray observations. In this article, results from the VERITAS observations of the Coma cluster of galaxies are reported, together with its constraints on cosmic-ray particle populations, magnetic fields, and dark matter in the cluster.

\section{Observations and Analysis}
The VERITAS gamma-ray detector \citep{article:Weekes_etal:2002} is an array of four 12 m diameter imaging atmospheric Cherenkov telescopes \citep[IACTs;][]{article:Holder_etal:2006} located at an altitude of 1268 m a.s.l. at the Fred Lawrence Whipple Observatory in southern Arizona (31$^{\circ}$~40\arcmin~30\arcsec~N, 110$^{\circ}$~57\arcmin~07\arcsec~W). Each of the telescopes is equipped with a 499 pixel camera covering a 3.5$^{\circ}$ field of view. The array, completed in the spring of 2007, is designed to detect gamma-ray emission from astrophysical objects in the energy range from 100 GeV to more than 30 TeV. The energy resolution is about 15\% and the angular resolution (68\% containment) is about 0.1$^{\circ}$ per event at 1 TeV. The sensitivity of the array allows for detection of a point source with a flux of 1\% of the steady Crab Nebula flux above 300 GeV at the $5\sigma$ level in under 30 hours.\footnote{The integral flux sensitivity above 300 GeV was improved by about 30\% with the move of one telescope in the summer of 2009. The move had no impact on the Coma cluster data set.}

@@

The Coma cluster is a very rich cluster of galaxies with many plausible sites for gamma-ray emission. The core region, the peripheral radio relic, as well as individual powerful cluster member galaxies. VERITAS has a large enough field of view to allow investigation of several of these scenarios. In this work the focus has been on the core region and three cluster members. The latter are treated as point sources in the analysis, whereas the core is treated as both a point source and a mildly extended source. The intrinsic source extension is assumed to be $0.2^{\circ}$ and $0.4^{\circ}$, based on the extension of the thermal soft X-ray emission from the core.

The ring-ring background model \citep{article:Aharonian_etal:2001} is used to estimate the background due to CRs. The total number of events in a given region of interest is then compared to the estimated background from the off-source region scaled by the ratio of the solid angles to produce a final excess. The VHE gamma-ray significance is then calculated using the method by \citet{article:LiMa:1983}.  Significance skymaps over the VERITAS field of view produced with a $0.2^{\circ}$ sampling radius are shown in Figure \ref{fig:skymaps} with overlaid X-ray and radio contours from the ROSAT all-sky survey \citep{article:BrielHenryBohringer:1992}  and GBT 1.4 GHz observations \citep{article:BrownRudnick:2010} respectively.

\section{Results}
No significant excess of VHE gamma-rays from the Coma cluster was detected with VERITAS, as illustrated by the $\theta^{2}$ distribution shown in  Figure \ref{fig:thetasq}. The $\theta^{2}$ distribution is a radial comparison of the ON-source event distribution with the estimated background. A 99\% confidence level upper limit is calculated for each region of interest using the method described by \citet{article:Rolke:2005} assuming a Gaussian background. Figure \ref{fig:sigdist} shows that the VERITAS data is well fit by a Gaussian with a mean close to zero and a standard deviation within a few percent of unity. The results are listed in Table \ref{table:results}.

@@

@@

\section{Conclusions}
VERITAS observed the Coma cluster of galaxies for a total of 18.6 hours of high-quality live time between March and May in 2008.

@@

\begin{figure*}
\begin{center}
\scalebox{0.45}{\plotone{f1a.eps}}
\scalebox{0.45}{\plotone{f1b.eps}}
\end{center}
\caption{\emph{Top}: Smoothed significance map of the Coma cluster calculated from the observed excess VHE gamma-ray events over a $4.5^{\circ}\times 4.5^{\circ}$ field of view. The excess counts were derived using a ring-ring background model \citep{article:Aharonian_etal:2001}. White contours show the X-ray counts per second in the 0.1 to 2.4 keV energy band from the ROSAT all-sky survey \citep{article:BrielHenryBohringer:1992}. \emph{Bottom}: Same as above but with overlaid contours from point source subtracted GBT radio observations at 1.4 GHz \citep{article:BrownRudnick:2010}. Shown are also the $0.2^{\circ}$ and $0.4^{\circ}$ radii (dashed cyan) considered for the extended source analyses presented here.}
\label{fig:skymaps}
\end{figure*}

\begin{figure*}
\begin{center}
\scalebox{1.0}{\plotone{f2.eps}}
\end{center}
\caption{$\theta^{2}$ distribution from VERITAS observations of the Coma cluster of galaxies. The points with error bars represent the ON-source data sample and and the filled area the background sample. Each bin represents an annulus around the Coma cluster core position. The data were derived from the ring-ring background model using a $0.2^{\circ}$ sampling radius.}
\label{fig:thetasq}
\end{figure*}

\begin{figure*}
\begin{center}
\scalebox{1.0}{\plotone{f3.eps}}
\end{center}
\caption{Distribution of significance for Figure \ref{fig:skymaps} and a sampling radius of $0.2^{\circ}$. The line is a Gaussian fit to the data, with mean close to zero and almost unit standard deviation.}
\label{fig:sigdist}
\end{figure*}

\begin{landscape}
\begin{deluxetable}{lccccccc}
\tablecolumns{7}
\tablewidth{0pc}
\tablecaption{Flux upper limits for regions of interest in the Coma cluster of galaxies.}
\tablehead{
\colhead{Source} &
\colhead{RA} &
\colhead{Dec} &
\colhead{RoI} &
\colhead{Excess} &
\colhead{Significance\tablenotemark{a}} &
\multicolumn{2}{c}{Flux U.L.\tablenotemark{b}}\\
\colhead{} & \colhead{[J2000]} & \colhead{[J2000]} & \colhead{[deg]} & \colhead{[counts]} & \colhead{[$\sigma$]} & \multicolumn{2}{c}{[$10^{-8}$ ph. m$^{-2}$ s$^{-1}$]}}
\startdata
Core & \RA{12}{59}{48.7} & \Dec{+27}{58}{50.0} & 0 & 17 & 0.84 & 2.62 & (1.3\%)\\
& & & 0.2 & -41 & -1.0 & 1.7 & (0.87\%)\\
& & & 0.4 & -26 & -0.30 & 4.5 & (2.3\%)\\
NGC 4889 & \RA{13}{00}{08.13} & \Dec{+27}{58}{37.03} & 0 & 3 & 0.14 & 2.09 & (1.1\%)\\
NGC 4874 & \RA{12}{59}{35.71} & \Dec{+27}{57}{33.37} & 0 & -14 & -0.71 & 1.42 & (0.72\%)\\
NGC 4921 & \RA{13}{01}{26.12} & \Dec{+27}{53}{09.59} & 0 & -4 & -0.23 & 1.89 & (0.96\%)
\enddata
\tablenotetext{a}{Statistical significance calculated according to \citet{article:LiMa:1983}}
\tablenotetext{b}{99\% confidence level upper limit calculated according to \citet{article:Rolke:2005} above an energy threshold of 220 GeV. Values in parentheses are the corresponding fluxes in percent of the steady Crab Nebula flux.}
\label{table:results}
\end{deluxetable}
\end{landscape}

>>>>

\altaffiltext{1}{Corresponding author: N. Karlsson, nkarlsson@physics.umn.edu}
\altaffiltext{2}{School of Physics and Astronomy, University of Minnesota, Minneapolis, MN 55455, USA}
\altaffiltext{3}{DESY, Platanenallee 6, 15738 Zeuthen, Germany}
\altaffiltext{4}{ Institut fu\"r Physik und Astronomie, Universit\"at Potsdam, 14476 Potsdam-Golm, Germany}
\altaffiltext{5}{Heidelberg Institute	for Theoretical	Studies, Schloss-Wolfsbrunnenweg 35, D-69118 Heidelberg, Germany}
\altaffiltext{6}{Canadian Institute for Theoretical Astrophysics, University of Toronto, Toronto, ON, M5S 3H8, Canada}
\altaffiltext{7}{University of California, Santa Barbara, CA 93106}

@@

The Coma cluster of galaxies (ACO 1656) is one of the most thouroughly studied clusters over all wavelengths \citep{article:Voges_etal:1999}. Located at a distance of about 100 Mpc \citep[$z=0.023$;][]{article:StrubleRood:1999}, it is one of the closest and also one of the most massive clusters \citep[$M \sim 10^{15}M_{\odot}$;][]{article:Smith:1983, article:Kubo_etal:2008}. It hosts both a giant radio halo \citep{article:Giovannini_etal:1993,article:Thierbach_etal:2003} and peripheral radio relic, which appears connected to the radio halo with a ``diffuse'' bridge \citep[see discussion in][]{article:BrownRudnick:2010}. It has been suggested that the relic is an infall shock. Extended soft X-ray (SXR) emission is evident from the ROSAT all-sky survey in the 0.1 to 2.4 keV band \citep{article:BrielHenryBohringer:1992} and observations with XMM-Newton \citep{article:Briel_etal:2001} revealed substructure in the X-ray halo. As such it is a natural candidate for gamma-ray observations. In this article, results from the VERITAS observations of the Coma cluster of galaxies are reported, together with the resulting constraints on cosmic-ray particle populations, magnetic fields, and dark matter in the cluster.

\section{VERITAS Observations and Analysis}
The VERITAS gamma-ray detector \citep{article:Weekes_etal:2002} is an array of four 12 m diameter imaging atmospheric Cherenkov telescopes \citep[IACTs;][]{article:Holder_etal:2006} located at an altitude of 1268 m a.s.l. at the Fred Lawrence Whipple Observatory in southern Arizona (31$^{\circ}$~40\arcmin~30\arcsec~N, 110$^{\circ}$~57\arcmin~07\arcsec~W). Each of the telescopes is equipped with a 499 pixel camera covering a 3.5$^{\circ}$ field of view. The array, completed in the spring of 2007, is designed to detect gamma-ray emission from astrophysical objects in the energy range from 100 GeV to more than 30 TeV. The energy resolution is about 15\% and the angular resolution (68\% containment) is about 0.1$^{\circ}$ per event at 1 TeV. The sensitivity of the array allows for detection of a point source with a flux of 1\% of the steady Crab Nebula flux above 300 GeV at the $5\sigma$ level in under 30 hours.\footnote{The integral flux sensitivity above 300 GeV was improved by about 30\% with the move of one telescope in the summer of 2009. The move had no impact on the Coma cluster data set.}

@@

The Coma cluster is a very rich cluster of galaxies with many plausible sites for gamma-ray emission. The core region, the peripheral radio relic, as well as individual powerful cluster member galaxies. VERITAS has a large enough field of view to allow investigation of several of these scenarios. In this work the focus has been on the core region and three cluster members. The latter are treated as point sources in the analysis, whereas the core is treated as both a point source and a mildly extended source. The intrinsic source extension is assumed to be $0.2^{\circ}$ and $0.4^{\circ}$, based on the extension of the thermal soft X-ray emission from the core. There is evidence of a recent merger event between the two central galaxies NGC 4889 and NGC 4874 \citep{article:Tribble:1993}. There is also evidence for an excess of non-thermal X-ray emission from these galaxies as well as the galaxy NGC 4921. Therefore, searches for point-like VHE gamma-ray emission have been conducted at the locations of these galaxies. The regions of interest considered in this work are summarized in Table \ref{table:roi}.

The ring background model \citep{article:Aharonian_etal:2001} is used to estimate the background due to CRs. The total number of events in a given region of interest is then compared to the estimated background from the off-source region scaled by the ratio of the solid angles to produce a final excess. The VHE gamma-ray significance is then calculated using the method by \citet{article:LiMa:1983}.  Significance skymaps over the VERITAS field of view produced with a $0.2^{\circ}$ sampling radius are shown in Figure \ref{fig:skymaps} with overlaid X-ray and radio contours from the ROSAT all-sky survey \citep{article:BrielHenryBohringer:1992}  and GBT 1.4 GHz observations \citep{article:BrownRudnick:2010} respectively.

%%\section{Fermi-LAT Observations and Analysis}

\section{Results}
No significant excess of VHE gamma-rays from the Coma cluster was detected with VERITAS, as illustrated by the $\theta^{2}$ distribution shown in  Figure \ref{fig:thetasq}. The $\theta^{2}$ distribution is a radial comparison of the ON-source event distribution with the estimated background. It extends out to a radius of $0.42^{\circ}$ to cover all regions of interest considered in this work. A 99\% confidence level upper limit is calculated for each region of interest using the method described by \citet{article:Rolke_etal:2005} assuming a Gaussian background. A lower bound of zero is imposed on the signal rate to handle the negative excesses seen for several of the regions of interest. Figure \ref{fig:sigdist} shows that the VERITAS data is well fit by a Gaussian with a mean close to zero and a standard deviation within a few percent of unity.

Table \ref{table:results} lists the upper limits for the selected regions of interest tabulated in Table \ref{table:roi}. These upper-limit calculations depend on the gamma-ray spectrum, which in this work is assumed to be a power law with a spectral index $\alpha=-2.1, -2.3$, and $-2.5$. A power-law spectrum with index $\alpha=-1.5$ is also considered for the dark matter constraints.

@@

\section{Constrainsts on Magnetic Fields}

@@

\section{Discussion}
VERITAS observed the Coma cluster of galaxies for a total of 18.6 hours of high-quality live time between March and May in 2008.

@@

\begin{figure*}
\begin{center}
\scalebox{0.45}{\plotone{f1a.eps}}
\scalebox{0.45}{\plotone{f1b.eps}}
\end{center}
\caption{\emph{Left}: Smoothed significance map of the Coma cluster calculated from the observed excess VHE gamma-ray events over a $4.5^{\circ}\times 4.5^{\circ}$ field of view. The excess counts were derived using a ring background model \citep{article:Aharonian_etal:2001}. White contours show the X-ray counts per second in the 0.1 to 2.4 keV energy band from the ROSAT all-sky survey \citep{article:BrielHenryBohringer:1992}. \emph{Right}: Same as above but with overlaid contours from point source subtracted GBT radio observations at 1.4 GHz \citep{article:BrownRudnick:2010}. Shown are also the $0.2^{\circ}$ and $0.4^{\circ}$ radii (dashed cyan) considered for the extended source analyses presented here.}
\label{fig:skymaps}
\end{figure*}

\begin{figure*}
\begin{center}
\scalebox{1.0}{\plotone{f2.eps}}
\end{center}
\caption{$\theta^{2}$ distribution from VERITAS observations of the Coma cluster of galaxies. The points with error bars represent the ON-source data sample and and the filled area the background sample. Each bin represents an annulus around the Coma cluster core position. The data were derived from the ring background model using a $0.2^{\circ}$ sampling radius.}
\label{fig:thetasq}
\end{figure*}

\begin{figure*}
\begin{center}
\scalebox{0.75}{\plotone{f3.eps}}
\end{center}
\caption{Distribution of significance for Figure \ref{fig:skymaps} and a sampling radius of $0.2^{\circ}$. The line is a Gaussian fit to the data, with mean close to zero and almost unit standard deviation.}
\label{fig:sigdist}
\end{figure*}

\begin{deluxetable}{lcc}
\tablecolumns{3}
\tablewidth{0pc}
\tablecaption{Regions of interest in the Coma cluster field of view. The cluster core is considered both as a point source and a modestly extended source. Three central galaxies are also considered in point-source searches. The choice is based on evidence of an excess of non-thermal X-ray emission \citep{article:Neumann_etal:2003} at the location of these galaxies.}
\tablehead{
\colhead{Source} &
\colhead{RA} &
\colhead{Dec}}
\startdata
Core & \RA{12}{59}{48.7} & \Dec{+27}{58}{50.0}\\
NGC 4889 & \RA{13}{00}{08.13} & \Dec{+27}{58}{37.03}\\
NGC 4874 & \RA{12}{59}{35.71} & \Dec{+27}{57}{33.37}\\
NGC 4921 & \RA{13}{01}{26.12} & \Dec{+27}{53}{09.59}\\
\enddata
\label{table:roi}
\end{deluxetable}

\begin{landscape}
\begin{deluxetable}{lcccccccccccc}
\tablecolumns{6}
\tablewidth{0pc}
\tablecaption{Flux upper limits for regions of interest in the Coma cluster of galaxies and surroundings.}
\tablehead{
\colhead{Source} & \colhead{RoI\tablenotemark{a} [deg]} & \colhead{$N_{S}$} & \colhead{S\tablenotemark{b} [$\sigma$]} &
\multicolumn{6}{c}{Flux U.L.\tablenotemark{c}} &
\multicolumn{3}{c}{Minimum gamma-ray flux\tablenotemark{d}}\\
\colhead{} & \colhead{} & \colhead{} & \colhead{} & \multicolumn{2}{c}{$\alpha=-2.1$} &
\multicolumn{2}{c}{$\alpha=-2.3$} &
\multicolumn{2}{c}{$\alpha=-2.5$} &
\colhead{$\alpha=-2.1$} &
\colhead{$\alpha=-2.3$} &
\colhead{$\alpha=-2.5$}}
\startdata
Core & 0 & 17 & 0.84 & 2.59 & (0.78\%) & 2.78 & (0.83\%) & 2.97 & (0.89\%) \\
& 0.2 & -41 & -1.0 & 1.96 & (0.59\%)  & 2.09 & (0.63\%) & 2.21 & (0.66\%) \\
& 0.4 & -26 & -0.30 & 4.44 & (1.3\%)  & 4.74 & (1.4\%) & 5.02 & (1.5\%)\\
NGC 4889 & 0 & 3 & 0.14 & - & - & 1.85 & (0.55\%)  & - & - \\
NGC 4874 & 0 & -14 & -0.71 & - & - &  1.51 & (0.45\%)  & - & - \\
NGC 4921 & 0 & -4 & -0.23 & - & - &  2.41 & (0.72\%)  & - & -
\enddata
\tablenotetext{a}{Intrinsic source radius (zero means point source), which is convolved with the analysis PSF.}
\tablenotetext{b}{Statistical significance calculated according to \citet{article:LiMa:1983}.}
\tablenotetext{c}{99\% confidence level upper limit in units of $10^{-8}$ ph. m$^{-2}$ s$^{-1}$ calculated according to \citet{article:Rolke_etal:2005} above an energy threshold of 220 GeV. Values in parentheses are the corresponding fluxes in percent of the steady Crab Nebula flux.}
\tablenotetext{d}{Minimum gamma-ray flux in units of $10^{-13}$ ph. cm$^{-2}$ s$^{-1}$ from the hadronic model described in the text.}
\label{table:results}
\end{deluxetable}
\end{landscape}
